---
title: "BSA6"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BSA6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=FALSE
)
```

# Setup

You are responsible for installing these packages prior to running this notebook

```{r setup}
library(BSA)
library(tidyverse)
library(foreach)
library(GenomicRanges)
library(BSgenome.CneoformansVarGrubiiKN99.NCBI.ASM221672v1)

kn99_genome = BSgenome.CneoformansVarGrubiiKN99.NCBI.ASM221672v1
```

# Read in data from VCF and parse into dataframe

```{r}
chr_map = tibble(seqnames = seqnames(kn99_genome), 
                 CHR=c(seq(1:14), "M"))

# the following file is from daniels original pipeline, which used a 
# hacked together set of aligners ngm + yaha
# system.file("BSA6.IGV.vcf.gz", package="BSA")
# The example uses the bwamem2 output from the current pipeline, and is  
# recommended
input_paths = list(
  bsa6_vcf = system.file('group_1_merged_dusted.recode.vcf.gz', package="BSA"),
  bsa6_meta = system.file("bsa6_samplesheet.rds", package="BSA")
)

meta_df = readRDS(input_paths$bsa6_meta)

tmp_dir = tempdir()
raw_samples_df = vcf_to_qtlseqr_table(input_paths$bsa6_vcf, "~/Desktop", 
                                      parent_ref_sample = 'KN99a', 
                                      parent_alt_sample = 'TDY1993',
                                      parent_filter = TRUE) 

samples_df = raw_samples_df %>% 
  filter(!sample %in% c("KN99a", "TDY1993")) %>%
  arrange(sample, CHR, POS) %>% 
  left_join(meta_df) %>%
  # note this!
  mutate(bulk = ifelse(cond == "inoculum", "low", "high"))
```

```{r}
summary(
  filter(samples_df, 
         Filt_Genotype == "Alternative", 
         genotype != 1) %>% 
          select(sample, Depth, RealDepth, genotype)
  )
```

```{r}

filter(samples_df, Filt_Genotype == "Alternative", genotype != 1) %>% 
  ggplot(aes(sample, RealDepth)) + 
  geom_boxplot() +
  ggtitle("Filt Genotype, VCF genotype mismatch. RealDepth Distributions")
```

```{r}
samples_df %>% 
  filter(sample == "P4L2") %>% 
  pull(RealDepth) %>%
  summary()
```

```{r}
samples_df %>%
  filter(!is.na(group)) %>%
ggplot(aes( sample, log2(Depth))) + 
    geom_violin(aes(fill=factor(cond))) + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
          text = element_text(size = 20)) +
  scale_y_continuous(limits = c(8,18), breaks = seq(8, 18, 1)) +
    facet_grid(group~day,scales = "free_x")
```


# Split the data into two groups

`oneMouse` are the samples passed through a single mouse. `sepMouse` are those 
that are passed through 5 separate mice.

```{r}
group_split = samples_df %>%
  # NOTE! I didn't have the WT in my samplesheet...
  filter(!is.na(group)) %>%
  group_by(group) %>%
  group_split()

names(group_split) = unlist(map(group_split, ~unique(pull(.,group))))

meta_df_group_split = meta_df %>%
  group_by(group) %>%
  group_split()

names(meta_df_group_split) = 
  unlist(map(meta_df_group_split, ~unique(pull(.,group))))
```

# Aggregate by pool and condition

In BSA6, we are comparing the effect of passing the variants through a single 
mouse vs separate mice. For each of these sets, we wish to examine data at
different levels of aggregation. Hence, here we aggregate first by pool to create 
the `summed_replicate` data, and then by condition to create the 
`summed_condition` data.

```{r}
# Sum over replicates
# For each given condition, the result is that there will be only one pool. So 
# if we had 2 replicates per pool, there will now be only one pool of YPD at 15 
# days. If there are 5 pools, each done in a different mouse, then there will be 
# 5 independent records (pools) of that YPD at 15 days
summed_replicates_split = map(group_split, 
                              collapse_bsa_data, 
                              c('CHR','POS','REF_Allele','ALT1',
                                'bulk','cond','pool','day'),
                              c('pool', 'cond', 'day'))

summed_replicates_meta = summed_replicates_split %>%
  map(select, sample, pool, cond, day) %>% 
  map(distinct, sample, .keep_all = TRUE)

# Sum over both replicates AND pools
# This creates a single record for a given condition, eg YPD at 15 days
summed_conditions_split = map(group_split, 
                              collapse_bsa_data, 
                              c('CHR','POS','REF_Allele','ALT1',
                                'bulk', 'cond','day'),
                              c('cond', 'day'))

summed_conditions_meta = summed_conditions_split %>%
  map(select, sample, cond, day, bulk) %>% 
  map(distinct, sample, .keep_all = TRUE)
```

# Ready the data for input into QTLseqR

We now need to transform the data into a format which the QTLseqR functions 
can consume. This is the purpose of `picker2`. Get more information on 
`picker2` and all of the functions in this vignette like so: `?picker2`.

```{r}
# this function will run the `picker2` function on the oneMouse and sepMouse 
# groups
run_picker2 = function(group_name,
                       metadata_split,
                       sample_df, 
                       grouping_var,
                       base_cond_in_each_group){

  experiment_crosses =
    sample_comparison_frame(metadata_split,
                            grouping_var = grouping_var,
                            base_cond_in_each_group = base_cond_in_each_group)

  out = foreach(
    row = iterators::iter(experiment_crosses, by='row'),
    .inorder = TRUE
  ) %do% {
      input_df = sample_df %>% 
        filter(sample %in% c(row[['lowBulk']], row[['highBulk']]))
      if (length(unique(input_df$bulk)) < 2){
        futile.logger::flog.info(paste0('Both samples ',
                                      row[['lowBulk']],
                                      ' and ',
                                      row[['highBulk']], 
                                      ' are `low` bulk. ',
                                      'Setting the second sample ',
                                      'to `high` for `picker2`'))
      input_df = input_df %>%
        mutate(bulk = ifelse(sample == row[['highBulk']], 'high', bulk))
      }
      tryCatch(picker2(input_df), 
               error = function(e){
                 futile.logger::flog.error(
                   paste0('row: ', row," ; error: ", e)
                 )
               })
  }
  names(out) = experiment_crosses$highBulk
  
  out
}

bsa_data_sets = list(
  
  # run picker2 on every record
  no_collapse         = map(names(group_split), 
                            ~run_picker2(
                              ., 
                              meta_df_group_split[[.]], 
                              group_split[[.]],
                              grouping_var = 'pool',
                              base_cond_in_each_group = TRUE)),
  
  # previously called summedReplicatesBSA
  replicates_collapse = map(names(summed_replicates_split),
                            ~run_picker2(
                              ., 
                              summed_replicates_meta[[.]],
                              summed_replicates_split[[.]],
                              grouping_var = 'pool',
                              base_cond_in_each_group = TRUE)),

  # previously called allPoolsInOneBSA
  condition_collapse  = map(names(summed_conditions_split),
                            ~run_picker2(
                              ., 
                              summed_conditions_meta[[.]],
                              summed_conditions_split[[.]],
                              grouping_var = 'day',
                              base_cond_in_each_group = FALSE))
  
  )

names(bsa_data_sets$no_collapse)         = names(group_split)

names(bsa_data_sets$replicates_collapse) = names(summed_replicates_split)

names(bsa_data_sets$condition_collapse)  = names(summed_conditions_split)
```

# Run the QTLSeqR functions to calculate the allele frequency metrics

Now that we have data which is aggregated at different levels, and since this is 
BSA6 this occurs for both the pooled samples (strains passed through separate 
mice) and oneMouse (all strains passed through a single mouse), we conduct this 
one both of those data sets (each with three levels of aggregate data).

```{r, echo = FALSE}
run_analyzer_on_bsa_set = function(bsa_group){
  outer_out = foreach(
    # the bsa_group is either oneMouse or sepMouse
    bsa_set = bsa_group,
    .inorder = TRUE
    ) %do% {
      inner_out = foreach(
        # each df is for a given sample
        df = bsa_set,
        .inorder = TRUE
      ) %do% { 
        analyzer(df[!is.nan(df$deltaSNP),], 
               minDepthPercentile = 0.1, 
               maxDepthPercentile = 0.995, 
               outlierFilt = "Hampel",
               filter_chr_list = c("CP022335.1", "G418", "NAT")) # omit mito and markers
      }
      names(inner_out) = names(bsa_set)
      inner_out
    }
  names(outer_out) = names(bsa_group)
  outer_out
}

filtered_bsa_data_sets = map(bsa_data_sets, run_analyzer_on_bsa_set)
```

# Plotting

## Create genome windows (tiles)

The table will look like this:

```raw
# A tibble: 18,921 × 6
   CHROM      start   end width strand CHR  
   <chr>      <int> <int> <int> <fct>  <chr>
 1 CP022321.1     1  1000  1000 *      1    
 2 CP022321.1  1001  2000  1000 *      1    
 3 CP022321.1  2001  3000  1000 *      1    
 4 CP022321.1  3001  4000  1000 *      1    
 5 CP022321.1  4001  5000  1000 *      1    
 6 CP022321.1  5001  6000  1000 *      1    
 7 CP022321.1  6001  7000  1000 *      1    
 8 CP022321.1  7001  8000  1000 *      1    
 9 CP022321.1  8001  9000  1000 *      1    
10 CP022321.1  9001 10000  1000 *      1  
```

```{r}
tiled_genome_df = 
  tileGenome(seqlengths(kn99_genome), 
             tilewidth = 1000, 
             cut.last.tile.in.chrom = TRUE) %>% 
  as_tibble() %>% 
  left_join(chr_map) %>%
  dplyr::rename(CHROM=seqnames)
```

#

```{r}
# a helper function whose purpose is to execute the `bin_variants` function
# on each sample in either oneMouse or pooled_mouse sets
bin_variants_by_group = function(set_name, bsa_set){
  out = foreach(
    sample_name = names(bsa_set),
    .inorder = TRUE
  ) %do% {
    df = bsa_set[[sample_name]]
    x = bin_variants(df, 
                 tiled_genome_df,
                 seqlengths(kn99_genome),
                 "CHROM") %>%
      mutate(group = set_name,
             sample = sample_name)
  }
  do.call('rbind', out)
}

bin_list = list(
  
  binned_filtered_summedReplicates = 
    map(
      names(filtered_bsa_data_sets$replicates_collapse), 
      ~bin_variants_by_group(
      .,filtered_bsa_data_sets$replicates_collapse[[.]])) %>%
  do.call('rbind',.),
  
  binned_filtered_summedConditions = 
    map(
      names(filtered_bsa_data_sets$condition_collapse), 
      ~bin_variants_by_group(
      .,filtered_bsa_data_sets$condition_collapse[[.]])) %>%
    do.call('rbind', .)
) 

graphing_df = bin_list %>%
  do.call('rbind', .) %>%
  mutate(collapse_level = 
           ifelse(str_detect(sample, "^\\d|^all"),
                  "replicate", "condition"))
```

```{r}
pool_min_max_df = graphing_df %>% 
  ungroup() %>%
  filter(collapse_level == "replicate") %>%
  separate(sample, c('pool', 'cond', 'day'), sep = "_", remove = FALSE) %>%
  select(-pool) %>%
  group_by(CHROM, tile, group, cond, day) %>% 
  summarize(pool_max_smoothed_allele_freq   = max(mean_smoothed_delta_snp, na.rm = TRUE),
            pool_min_smoothed_allele_freq   = min(mean_smoothed_delta_snp, na.rm = TRUE),
            pool_max_unsmoothed_allele_freq = max(mean_deltaSNP, na.rm = TRUE),
            pool_min_unsmoothed_allele_freq = min(mean_deltaSNP, na.rm = TRUE),
            .groups = 'keep') %>%
  separate(tile, c("binFloor","binCeiling"), sep = ",") %>%
  ungroup() %>%
  pivot_longer(starts_with("pool")) %>%
  mutate(min_max = paste0("pool_allele_freq_", str_extract(name, "max|min")),
         smoothed = ifelse(str_extract(name, "smoothed|unsmoothed") == 'smoothed', TRUE, FALSE)) %>%
  select(-name) %>%
  pivot_wider(names_from = min_max, values_from = value) %>%
  arrange(group, cond, day, CHROM, binFloor) %>%
  arrange(smoothed) %>%
  select(group,cond,day, CHROM,binFloor,binCeiling,smoothed, 
         pool_allele_freq_min, pool_allele_freq_max) %>%
  mutate(binFloor = as.numeric(binFloor), binCeiling = as.numeric(binCeiling))

condition_collapsed_df = graphing_df %>%
  ungroup() %>%
  filter(collapse_level == "condition") %>% 
  separate(sample, c('cond', 'day'), sep = "_", remove = FALSE) %>%
  select(group,cond,day,significance, CHROM, binFloor, binCeiling, binMiddle, mean_deltaSNP, mean_smoothed_delta_snp) %>%
  pivot_longer(c(mean_deltaSNP, mean_smoothed_delta_snp)) %>%
  mutate(smoothed = ifelse(str_detect(name, "smoothed"), TRUE, FALSE)) %>%
  dplyr::rename(condition_allele_freq=value) %>%
  select(-name) %>%
  arrange(group, cond, day, CHROM, binFloor) %>%
  arrange(smoothed) %>%
  select(group,cond,day,significance, CHROM,binFloor, 
         binCeiling,smoothed, condition_allele_freq)

x = pool_min_max_df %>%  
  mutate(binFloor = as.numeric(binFloor), binCeiling = as.numeric(binCeiling)) %>%
  left_join(condition_collapsed_df) %>%
  filter(complete.cases(.)) %>%
  mutate(binMiddle = (binFloor+binCeiling )/ 2) %>%
    select(group,cond,day,significance, CHROM,binFloor, 
         binCeiling,binMiddle,smoothed,pool_allele_freq_min, 
         pool_allele_freq_max, condition_allele_freq)


# TODO this needs to be checked
#The SNPindex.LOW should be the same for YPD and Lungs....
MaxAndMins = filtered_bsa_data_sets$condition_collapse$sepMouse$ypd_1 %>% 
  select(c(CHROM,POS,REF,ALT, SNPindex.LOW)) %>%
  mutate(min_SNPchange = -1*SNPindex.LOW, max_SNPchange = 1-SNPindex.LOW) %>% 
  select(-SNPindex.LOW)

#This mutation was erased in the addition of the drug marker in the C8 parent (It was in the arm).
MaxAndMins=MaxAndMins %>% 
  filter(!(CHROM=="chr2" & 
             (POS==283162 | POS==283652| 
                (POS>466000 & POS<467000)))) 
```

```{r}
chr_name = "CP022322.1"

x %>%
  filter(CHROM==chr_name, cond %in% c("lung", "ypd"), smoothed == TRUE) %>%
  unite("combine", cond,significance, remove = FALSE) %>%
  ggplot(aes(binMiddle/1000, condition_allele_freq)) +
    geom_line(data = filter(MaxAndMins, CHROM == chr_name), aes(POS/1000, min_SNPchange)) +
    geom_line(data = filter(MaxAndMins, CHROM == chr_name), aes(POS/1000, max_SNPchange)) +
    geom_pointrange(aes(ymin=pool_allele_freq_min,
                        ymax=pool_allele_freq_max,
                        x=binMiddle/1000,
                        y=condition_allele_freq,
                        size=combine,
                        colour=combine,
                        alpha=combine,
                        order="combine"), na.rm = TRUE) +
    scale_size_manual(name="Condition and significance",
                      labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
                      values = c(0.1,0.1,0.4, 0.4), drop=F)+
    scale_colour_manual(name="Condition and significance",
                        labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
                        values = c('lung_0'="red",'ypd_0'="black",
                                   'lung_1'="red",'lung_1'="black"), drop=F)+
    scale_alpha_manual(name="Condition and significance",
                       labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
                       values = c(0.95,0.95,1, 1), drop=F)+
    labs(title=paste0("Chromosome ", chr_name), x=paste0("Position on chromosome ",chr_name," (kb)"),y="Change in allele frequency (BSA)", color="Pool")+
    geom_hline(yintercept = c(0), color="black", size=0.6)+
    scale_x_continuous(breaks = seq(from=0, to=2400, by=300), limits=c(0, 2400),expand = c(0, 0))+
    scale_y_continuous(breaks = c(-1,-0.75,-0.5, -0.25, 0,0.25,0.5,0.75,1), limits = c(-1.05,1.05), expand = c(0,0))+
    theme(#legend.title = element_text(size=18),
      legend.position = "right",
      plot.title = element_blank(),
      legend.text = element_text(size=16),
      axis.title=element_text(size=32),
      axis.title.y =element_text(margin = margin(r=25, t=0, l=5, b=0)),
      axis.title.x =element_text(margin = margin(r=0, t=25, l=0, b=5)),
      axis.text=element_text(size = 28),
      axis.ticks.length = unit(0.3,"cm"),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      panel.background = element_rect(fill = "white",colour = "black", size=1),
      text=element_text(family="sans"))+
    facet_grid(day~group)
```

```{r}
x %>%
  filter(CHROM==chr_name, cond %in% c("lung", "ypd"), smoothed == TRUE) %>%
  unite("combine", cond,significance, remove = FALSE) %>%
  ggplot(aes(binMiddle/1000, condition_allele_freq)) +
    geom_line(data = filter(MaxAndMins, CHROM == chr_name), aes(POS/1000, min_SNPchange)) +
    geom_line(data = filter(MaxAndMins, CHROM == chr_name), aes(POS/1000, max_SNPchange)) +
    geom_pointrange(aes(ymin=pool_allele_freq_min,
                        ymax=pool_allele_freq_max,
                        x=binMiddle/1000,
                        y=condition_allele_freq,
                        size=combine,
                        colour=group,
                        alpha=combine,
                        order="combine"), na.rm = TRUE) +
    scale_size_manual(name="Condition and significance",
                      labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
                      values = c(0.1,0.1,0.4, 0.4), drop=F)+
    # scale_colour_manual(name="Condition and significance",
    #                     labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
    #                     values = c('lung_0'="red",'ypd_0'="black",
    #                                'lung_1'="red",'lung_1'="black"), drop=F)+
    scale_alpha_manual(name="Condition and significance",
                       labels=c("Lung, not significant","YPD, not significant","Lung, significant",  "YPD, significant"),
                       values = c(0.95,0.95,1, 1), drop=F)+
    labs(title=paste0("Chromosome ", chr_name), x=paste0("Position on chromosome ",chr_name," (kb)"),y="Change in allele frequency (BSA)", color="Pool")+
    geom_hline(yintercept = c(0), color="black", size=0.6)+
    scale_x_continuous(breaks = seq(from=0, to=2400, by=300), limits=c(0, 2400),expand = c(0, 0))+
    scale_y_continuous(breaks = c(-1,-0.75,-0.5, -0.25, 0,0.25,0.5,0.75,1), limits = c(-1.05,1.05), expand = c(0,0))+
    theme(#legend.title = element_text(size=18),
      legend.position = "right",
      plot.title = element_blank(),
      legend.text = element_text(size=16),
      axis.title=element_text(size=32),
      axis.title.y =element_text(margin = margin(r=25, t=0, l=5, b=0)),
      axis.title.x =element_text(margin = margin(r=0, t=25, l=0, b=5)),
      axis.text=element_text(size = 28),
      axis.ticks.length = unit(0.3,"cm"),
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      panel.background = element_rect(fill = "white",colour = "black", size=1),
      text=element_text(family="sans"))+
    facet_grid(~day)

```









