---
title: "bsa6_eda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bsa6_eda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# reduction in code from 8129 characters to 4860 characters. That includes 
# comments, spaces, headings, etc

```{r setup}
library(BSA)
library(tidyverse)
library(foreach)
library(here)
library(BSgenome.CneoformansVarGrubiiKN99.NCBI.ASM221672v1)

kn99_genome = BSgenome.CneoformansVarGrubiiKN99.NCBI.ASM221672v1
```

# Read in data from VCF and parse into dataframe

```{r}
chr_map = tibble(seqnames = seqnames(kn99_genome), 
                 CHR=c(seq(1:14), "M"))

input_paths = list(
  bsa6_vcf = system.file("BSA6.IGV.vcf.gz", package="BSA"),
  bsa6_meta = system.file("bsa6_samplesheet.rds", package="BSA")
)

meta_df = readRDS(input_paths$bsa6_meta)

tmp_dir = tempdir()
samples_df = vcf_to_qtlseqr_table(input_paths$bsa6_vcf, "~/Desktop")  %>% 
  left_join(chr_map) %>% 
  select(c("seqnames", colnames(samples_df)[2:length(samples_df)])) %>%
  dplyr::rename(CHR=seqnames) %>% 
  filter(!is.na(CHR)) %>%
  arrange(sample, CHR, POS)
```

```{r}
unique(samples_df$CHR)
```


# Join the VCF data with the sample metadata

Add column `bulk`. This refers to the structure of a BSA experiment. In this 
application, "low bulk" are the inocolum samples while high bulk are all other 
samples.

```{r}
samples_df_with_meta = samples_df %>% 
  left_join(meta_df) %>%
  # note this!
  mutate(bulk = ifelse(cond == "inoculum", "low", "high"))
```

# Split the data into two groups

`oneMouse` are the samples passed through a single mouse. `sepMouse` are those 
that are passed through 5 separate mice.

```{r}
group_split = samples_df_with_meta %>%
  # NOTE! I didn't have the WT in my samplesheet...
  filter(!is.na(group)) %>%
  group_by(group) %>%
  group_split()

names(group_split) = unlist(map(group_split, ~unique(pull(.,group))))

meta_df_split = meta_df %>%
  group_by(group) %>%
  group_split()

names(meta_df_split) = unlist(map(meta_df_split, ~unique(pull(.,group))))
```

# BSA on the replicates

```{r}
# this function will run the `picker2` function on the oneMouse and sepMouse 
# groups
replicate_bsa_by_group = function(group_name,
                                  metadata_split, 
                                  sample_df, 
                                  grouping_var,
                                  base_cond_in_each_group){

  experiment_crosses =
    sample_comparison_frame(metadata_split,
                            grouping_var = grouping_var,
                            base_cond_in_each_group = base_cond_in_each_group)

  out = foreach(
    row = iterators::iter(experiment_crosses, by='row'),
    .inorder = TRUE
  ) %do% {
    picker2(sample_df, 
            low_bulk_sample = row[['lowBulk']], 
            high_bulk_sample = row[['highBulk']])
  }
  names(out) = experiment_crosses$highBulk
  
  out
}
```

# Create QTLseqR input using picker2 on all samples

```{R}
replicatesBSA = map(names(group_split),
                    ~replicate_bsa_by_group(.,meta_df_split[[.]],
                                            group_split[[.]],
                                            grouping_var = 'pool',
                                            base_cond_in_each_group = TRUE))
names(replicatesBSA) = names(group_split)
```

# Collapse replicates and re-run picker2

```{r}

sum_over_replicates = function(df_with_meta){
  df_with_meta %>% 
  group_by(CHR,POS,REF_Allele,ALT1,bulk,cond,pool,day) %>% 
  summarize(RealDepth    = sum(RealDepth),
            Depth        = sum(Depth),
            Reference    = sum(Reference),
            Alternative1 = sum(Alternative1),
            .groups = 'keep') %>% 
  unite("sample", pool, cond, day, remove = FALSE) %>%
  group_by(sample) %>% 
  arrange(CHR, POS, REF_Allele, ALT1) %>%
  ungroup()
}

summed_replicates_split = map(group_split, sum_over_replicates)

summed_replicates_meta = summed_replicates_split %>%
  map(select, sample, pool, cond, day) %>% 
  map(distinct, sample, .keep_all = TRUE)

summedReplicatesBSA = map(names(summed_replicates_split),
                    ~replicate_bsa_by_group(.,summed_replicates_meta[[.]],
                                            summed_replicates_split[[.]],
                                            grouping_var = 'pool',
                                            base_cond_in_each_group = TRUE))

names(summedReplicatesBSA) = names(summed_replicates_split)

```

# Collapse conditions and re-run picker2
This collapsed both pool and replicate

```{r}

sum_over_conditions = function(df_with_meta){
  df_with_meta %>% 
  group_by(CHR,POS,REF_Allele,ALT1,bulk,cond, day) %>% 
  summarize(RealDepth    = sum(RealDepth),
            Depth        = sum(Depth),
            Reference    = sum(Reference),
            Alternative1 = sum(Alternative1),
            .groups = 'keep') %>% 
  unite("sample", cond, day, remove = FALSE) %>%
  group_by(sample) %>% 
  arrange(CHR, POS, REF_Allele, ALT1) %>%
  ungroup()
}

summed_conditions_split = map(group_split, sum_over_conditions)

summed_conditions_meta = summed_conditions_split %>%
  map(select, sample, cond, day, bulk) %>% 
  map(distinct, sample, .keep_all = TRUE)

allPoolsInOneBSA = map(names(summed_conditions_split),
                       ~replicate_bsa_by_group(.,summed_conditions_meta[[.]],
                                            summed_conditions_split[[.]],
                                            grouping_var = 'day',
                                            base_cond_in_each_group = FALSE))

names(allPoolsInOneBSA) = names(summed_conditions_split)
```

```{r}
filter_bsa_set = function(bsa_set){
  out = foreach(
    df = bsa_set,
    .inorder = TRUE
    ) %do% {
      analyzer(df[!is.nan(df$deltaSNP),], 
               minDepthPercentile = 0.1, 
               maxDepthPercentile = 0.995, 
               outlierFilt = "Hampel",
               filter_chr_list = c("CP022335.1", "G418", "NAT"))
    }
  names(out) = names(bsa_set)
  out
}

filtered_summedReplicatesBSA = map(summedReplicatesBSA, filter_bsa_set)

filtered_allPoolsInOneBSA = map(allPoolsInOneBSA, filter_bsa_set)

```

# Below is the plotting

```{r}
### PLOTS

chosenSize <- 5000
# Pools <- list()
# allPools <- list()
# 
# ptm <- proc.time()
# no_cores <- detectCores()-1 # length(names(pools_filteredBSA)) 
# cl <- makeCluster(no_cores)
# 
# # export data, functions and variables to cluster
# clusterExport(cl, c("filtered_summedReplicatesBSA",
#                     "binner2", 
#                     "Pools", 
#                     "allPools", 
#                     "chosenSize"))
# 
# # Calling the binner2 in parallel
# Pools <- parLapply(cl, names(pools_filteredBSA),
#   function(x) {
#     # make sure this is documented
#     binner2(pools_filteredBSA[[grep(x, names(pools_filteredBSA))]], bin.size = chosenSize)
#   }
# )
# stopCluster(cl) # Closing the parallel cluster
# proc.time() - ptm

bin_variants = function(variant_metrics_df, 
                        tiled_genome_df,
                        chr_seqlengths,
                        chr_colname,
                        qvalue_lower_thres = 0.1,
                        qvalue_upper_thres = 0.5){
  
  if(!chr_colname %in% colnames(variant_metrics_df)){
    stop(sprintf("%s not in variant_metrics_df colnames", chr_colname))
  } else if(!chr_colname %in% colnames(tiled_genome_df)){
    stop(sprintf("%s not in tiled_genome_df colnames", chr_colname))
  }
  
  final_col_order = c(chr_colname, 'binFloor', 'binCeiling', 
                      'binMiddle', 'mean_deltaSNP', 'smoothedDeltaSNP',
                      'mean_qvalue', 'mut_perBin', 'bin_size', 'singificance')
  
  variant_metrics_chr_split = variant_metrics_df %>%
  group_by(!!rlang::sym(chr_colname)) %>%
    group_split()
  
  names(variant_metrics_chr_split) = unlist(
    map(variant_metrics_chr_split, ~unique(pull(.,chr_colname)))
  )
  
  tiled_genome_chr_split = tiled_genome_df %>%
    group_by(!!rlang::sym(chr_colname)) %>%
    group_split()
  
  names(tiled_genome_chr_split) = unlist(
    map(tiled_genome_chr_split, ~unique(pull(.,chr_colname)))
  )
  
  out = map(names(variant_metrics_chr_split), 
        ~tile_metrics(variant_metrics_chr_split[[.]], 
                      tiled_genome_chr_split[[.]],
                      chr_seqlengths[[.]])) %>%
    do.call('rbind', .) %>%
    group_by(!!rlang::sym(chr_colname), tile) %>%
    summarize(mean_deltaSNP = mean(deltaSNP, na.rm=TRUE),
              smoothedDeltaSNP   = mean(tricubeDeltaSNP, na.rm = TRUE),
              mean_qvalue        = mean(qvalue, na.rm = TRUE)
              qval_below_thres_n = sum(qvalue < qvalue_lower_thres),
              qval_above_thres_n = sum(qvalue >= qvalue_upper_thres),
              mut_perBin         = n(),
              .groups = 'keep') %>%
    mutate(significance = qval_below_thres_n / qval_above_thres_n,
           tile = str_remove_all(tile, '\\[|\\]|\\(|\\)')) %>%
    separate(tile, c('binFloor', 'binCeiling'), sep=",") %>%
    mutate(binFloor = as.integer(binFloor),
           binCeiling = as.integer(binCeiling)) 
  
  if(length(setdiff(final_col_order, colnames(out))) > 0){
    stop(paste0(
          sprintf("The following columns are expected in the output: %s.\n",
                  paste(final_col_order, collapse = ",")), 
          sprintf("But the output columns do not possess one or more: %s",
                  paste(colnames(out), collapse=","))) )
  } else{
    out %>%
      select(all_of(final_col_order))
  }

    
  
}

tile_metrics = function(metrics_df, 
                        genome_tile_df, chr_seqlength,
                        cut_col = "POS", 
                        left_inclusive = TRUE, 
                        right_inclusive = FALSE){
  
  metrics_df %>% 
    mutate(tile = 
             cut(!!rlang::sym(cut_col), 
                 c(genome_tile_df$start, chr_seqlength + 1), 
                 include.lowest = left_inclusive, 
                 right = right_inclusive))

}

tiled_genome_df = tiled_genome_df %>%
  dplyr::rename(CHROM=seqnames) 
  

x = bin_variants(filtered_allPoolsInOneBSA$oneMouse$brain_15, 
                 tiled_genome_df,
                 seqlengths(kn99_genome),
                 "CHROM")

x  %>%
    summarize(deltaSNP = mean(deltaSNP, na.rm=TRUE),
              tricubeDeltaSNP = mean(tricubeDeltaSNP, na.rm = TRUE)) %>%
  mutate(qval_below_thres_n = count(qvalue < .1))

%>%
    mutate(significance = qval_below_thres_n / qval_above_thres_n)

binner2_by_group = function(bsa_set){
  foreach(
    df = bsa_set,
    .inorder = TRUE
  ) %do% {
    binner2(df, bin.size = chosenSize)
  }
}

x = binner2(filtered_allPoolsInOneBSA$sepMouse$lung_8)

# binned_filtered_summedReplicatesBSA = map(summedReplicatesBSA, binner2_by_group)



```

























